---
title: Solution to mystery analysis
description: Remember, the most important aspect of this exercise is the workflow that you followed, the decisions you made and the support (evidence) used to support those decisions.
author: 
  - name: Eric Rexstad
date: 02-28-2025
image: https://images.unsplash.com/photo-1443521156453-f82c842b6e7f?ixlib=rb-4.0.3&ixid=MnwxMjA3fDB8MHxjb2xsZWN0aW9uLXBhZ2V8NjB8NDcxMTQ0NHx8ZW58MHx8fHw%3D&auto=format&fit=crop&w=500&q=60
draft: true
---

```{r}
#| label = "setup",
#| include = FALSE
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, comment = NA)
library(kableExtra)
library(vioplot)
library(Distance)
```

:::{.callout-tip}
# Demonstration

Start-to-finish analysis of mystery data set
:::

# Data import

No drama to bring the labelled data frame into our **R** session.

```{r}
#| label = "read"
mydata <- read.csv("mystery.csv")
```

# Exploratory data analysis

Number of strata, number of transects, number of detections, detections by sex

```{r}
#| label = "samples",
#| results = "hold"
cat("Num strata= ", length(unique(mydata$Region.Label)))
cat("\nNum transects= ", length(unique(mydata$Sample.Label)))
cat("\nNum detects= ", sum(!is.na(mydata$distance)), "\n")
table(mydata$sex)
```

# Distance distribution

```{r}
#| label = "disdis",
#| results = "hold"
hist(mydata$distance, main="All detections", breaks=seq(0,80,length=17),
     xlab="Perpendicular distances (m)")
vioplot(mydata$distance[mydata$sex=="male"],
        mydata$distance[mydata$sex=="female"])
```

# Truncation decision

:::{.callout-hint}
# A *rule of thumb* (Lecture 2, slide 21)

> Can also use estimated values of g(x) from fitted model as truncation criterion; truncate at w when g(w)=0.15

:::

I'm reasonably content to stick with this rule of thumb, could slice out one more data point, if I truncated at 50m.

```{r}
#| label = "trunc",
#| message = FALSE
myunits <- convert_units("meter", "kilometer", "square kilometer")
first <- ds(mydata, convert_units = myunits)
```

```{r}
#| label = "plottrunc"
plot(first, breaks=seq(0,80,length=17))
abline(h=0.15, col="red", lty=3)
text(x=50, y=0.2, "Pr(detect)=0.15", cex=.8)
truncate <- 50
```

# Key function decision

```{r}
#| message = FALSE
unicos <- ds(mydata, key="unif", adj="cos", truncation = truncate, convert_units = myunits)
hn <- ds(mydata, key="hn", adj="cos", truncation = truncate, convert_units = myunits)
hr <- ds(mydata, key="hr", adj="cos", truncation = truncate, convert_units = myunits)
```

Notice, adjustment terms did not survive the "within key function family" selection.

```{r}
kable(summarize_ds_models(unicos, hn, hr, output="plain"), digits=3, row.names = FALSE,
                    caption="Key function models at 50m: do they fit?")  %>%
      kable_styling(full_width = F) %>%
      column_spec(4, background="yellow", width="8em")
```

# What about the covariate?

Recall, there are only 4 detections of females (10\% of data set).  Possibly, that few detections might not have a strong influence on the parameter estimates of the detection function.  Nevertheless, apply the covariate to our models at our truncation distance:

```{r}
#| label = "withcov",
#| message = FALSE
hn.sex <- ds(mydata, key="hn", truncation = truncate, convert_units = myunits,
             formula=~sex)
hr.sex <- ds(mydata, key="hr", truncation = truncate, convert_units = myunits,
             formula=~sex)
# unicos.sex <- ds(mydata, key="unif",  truncation = truncate, convert_units = myunits,
# formula=~sex)
```

`r knitr::asis_output("\U1F605")` uniform key function does not have a scale parameter $\sigma$, hence cannot be fitted.  That reduces the number of models of interest to us.  We are left with the half normal and hazard rate keys, with and without the sex covariate and the uniform cosine that cannot support a covariate.  

Do the models with covariates fit the data?  I expect they should because adding more parameters to a model ought to improve fit.  The half normal and hazard already fit the data without the covariate.

```{r}
#| label = "gof"
hn.sex.fit <- gof_ds(hn.sex, plot = FALSE, chisq = FALSE)$dsgof$CvM$p
hr.sex.fit <- gof_ds(hr.sex, plot = FALSE, chisq = FALSE)$dsgof$CvM$p
covfits <- data.frame(modname=c("HN sex", "HR sex"), 
                      CvMP=round(c(hn.sex.fit, hr.sex.fit),2))
kable(covfits) %>%
  kable_styling(full_width=FALSE, bootstrap_options="condensed")
```

## AIC among remaining five competitors

```{r}
#| label = "aic"
AIC(unicos, hn, hr, hn.sex, hr.sex)
```

What emerges?  Half normal and uniform with cosine adjustment have very similar shapes and are quite similar models.  The hazard rate key without a covariate seems out of contention.  What about the similarity in $\hat{P_a}$ between competing models?

```{r}
#| label = "lasttable"
kable(summarize_ds_models(unicos, hn,hr, hn.sex, hr.sex, output="plain"), digits=3, row.names = FALSE,
                    caption="Five competing models with truncation at 50m")  %>%
      kable_styling(full_width = F) %>%
      column_spec(5, background="yellow", width="8em")
```

Models without sex as a covariate estimate a larger $\hat{P_a}$, so even with only 4 female detections, those detections do exert an influence upon the shape of the estimated detection function and consequently upon $\hat{P_a}$.

# Are estimated $\hat{P_a}(z_i)$ too small?

```{r}
#| label = "pdist"
kable(p_dist_table(hn.sex, bins=seq(0,0.8,0.1), proportion=TRUE), digits=3,
      caption="Estimated detection probabilities from HN with sex covariate.") %>%
  kable_styling(full_width=FALSE, bootstrap_options="condensed")
```
This distribution violates the guideline that <5\% of $\hat{P_a}$ should be less than 0.2.  But I'm willing to overlook that.  

# Estimated detection function

```{r}
#| label = "bestdet"
summary(hn.sex$ddf)
```
```{r}
#| label = "plotfn"
plot(hn.sex, main="Half normal with sex covariate")
add_df_covar_line(hn.sex, mydata, lty=1, lwd=2, col=c("red", "blue"))
legend("topright", 
       c("Average", "Males", "Females"),
       col=c("black", "red", "blue"),
       lty=1)
```

Note the average (across sexes) detection probability curve is displaced toward the males that represent the largest proportion of the detections.

# Estimated abundance

## Best model, Half normal with sex covariate
```{r}
print(hn.sex$dht$individual$N)
```

If we wanted to employ the *gold standard* in precision estimation, we would apply a bootstrap

```{r}
#| label = "booter",
#| results = "hide"
bootout <- bootdht(hn.sex, flatfile=mydata, summary_fun = bootdht_Nhat_summarize,
                   nboot=500, convert_units = myunits)
```

```{r}
#| label = "plotit"
hist(bootout$Nhat, breaks = 20, 
     main="Distribution of bootstrap replicates", xlab="Abundance estimate")
mybounds <- round(quantile(bootout$Nhat, c(0.025, 0.975),na.rm=TRUE))
abline(v=mybounds, lwd=2, lty=3)
```

Confidence interval bounds from bootstrap are (`r mybounds`), somewhat wider than the analytical confidence interval bounds specified above.

## Closest AIC score competitor, Hazard rate with sex covariate
```{r}
print(hr.sex$dht$individual$N)
```

Notice the price, in terms of precision, paid for the extra parameter estimated in the hazard rate model compared to the half normal model.

```{r}
#| fig.width = 7,
#| fig.cap = "Point and 95% interval estimates based upon truncation distance of 50m.",
#| echo = FALSE
x <- rbind(hn.sex$dht$individuals$N, hr.sex$dht$individuals$N,
           unicos$dht$individuals$N, hn$dht$individuals$N, 
           hr$dht$individuals$N)
x$Label <- c("Hn~sex", "Hr~sex", "Unicos", "Hn", "Hr")
plot(1:5, x$Est, type="p", ylim=c(80,550), xaxt="n", xlab="Model", 
     ylab="Abundance estimate", pch=20, cex=2, main="Mystery data analysis",
     sub="Truth horizontal dashed line", font.main=4, font.sub=3)
axis(1, at=1:5, labels=x$Label)
nests <- seq(dim(x)[1])
segments(nests, x$lcl[nests], nests, x$ucl[nests], lwd=2)
abline(h=180, lwd=3, lty=3)
```
